---
title: "BIS 15L Midterm 2"
author: "Benjamin Prieto"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 10 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! Your plots should use consistent aesthetics throughout.  

This exam is due by **12:00p on Tuesday, February 22**.  

## Gapminder
For this assignment, we are going to use data from  [gapminder](https://www.gapminder.org/). Gapminder includes information about economics, population, social issues, and life expectancy from countries all over the world. We will use three data sets, so please load all three as separate objects.    

1. population_total.csv  
2. income_per_person_gdppercapita_ppp_inflation_adjusted.csv  
3. life_expectancy_years.csv  

```{r}
population <- read.csv("data/population_total.csv")
economy <- read.csv("data/income_per_person_gdppercapita_ppp_inflation_adjusted.csv")
life_exp <- read.csv("data/life_expectancy_years.csv")
```
<br/>
**1. (3 points) Once you have an idea of the structure of the data, please make each data set tidy (hint: think back to pivots) and store them as new objects. You will need both the original (wide) and tidy (long) data!**  

Setup
```{r}
library("tidyverse") #load packages
library("ggplot2")
library("ggthemes")
```

```{r}
options(scipen=999) #cancel scientific notation
```
<br/>
`population` Overview and Tidying
```{r}
glimpse(population)
```

```{r}
population_long <- population %>% 
  pivot_longer(cols = contains("X"), names_to = "year", values_to = "population") %>% 
  separate(year, into = c("var1", "year"), sep = "X") %>% 
  select(-var1)
population_long
```
<br/>
`economy` Overview and Tidying
```{r}
glimpse(economy)
```

```{r}
economy_long <- economy %>% 
  pivot_longer(cols = contains("X"), names_to = "year", values_to = "income_per_person_gdppercapita_ppp_inflation_adjusted") %>% 
  separate(year, into = c("var1", "year"), sep = "X") %>% 
  select(-var1)
economy_long
```
<br/>
`life_exp` Overview and Tidying
```{r}
glimpse(life_exp)
```

```{r}
life_exp_long <- life_exp %>% 
  pivot_longer(cols = contains("X"), names_to = "year", values_to = "life_expectancy_years") %>% 
  separate(year, into = c("var1", "year"), sep = "X") %>% 
  select(-var1)
life_exp_long
```
<br/>
<br/>
**2. (1 point) How many different countries are represented in the data? Provide the total number and their names. Since each data set includes different numbers of countries, you will need to do this for each one.**  

`population` 
```{r}
population %>% 
  summarize(n_countries = n_distinct(country))
```

```{r}
population %>% 
  select(country)
```
<br/>
`economy`
```{r}
economy %>% 
  summarize(n_countries = n_distinct(country))
```

```{r}
economy %>% 
  select(country)
```
<br/>
`life_exp`
```{r}
life_exp %>% 
  summarize(n_countries = n_distinct(country))
```

```{r}
life_exp %>% 
  select(country)
```

## Life Expectancy  

**3. (2 points) Let's limit the data to 100 years (1920-2020). For these years, which country has the highest average life expectancy? How about the lowest average life expectancy?**  

Highest Average Life Expectancy (1920-2020)
```{r}
life_exp_long %>% 
  filter(between(year, 1920, 2020)) %>% 
  group_by(country) %>% 
  summarize(avg_life_expectancy_years = mean(life_expectancy_years, na.rm = T)) %>% 
  top_n(n = 1)
```
<br/>
Lowest Average Life Expectancy (1920-2020)
```{r}
life_exp_long %>% 
  filter(between(year, 1920, 2020)) %>% 
  group_by(country) %>% 
  summarize(avg_life_expectancy_years = mean(life_expectancy_years, na.rm = T)) %>% 
  arrange(avg_life_expectancy_years) %>% 
  head(n = 1)
```
<br/>
<br/>
**4. (3 points) Although we can see which country has the highest life expectancy for the past 100 years, we don't know which countries have changed the most. What are the top 5 countries that have experienced the biggest improvement in life expectancy between 1920-2020?** 

```{r}
life_exp_long %>% 
  filter(year == 1920 | year == 2020) %>% 
  group_by(country) %>% 
  summarize(life_expectancy_change_years = max(life_expectancy_years) - min(life_expectancy_years)) %>% 
  top_n(n = 5) %>% 
  arrange(desc(life_expectancy_change_years))
```
<br/>
<br/>
**5. (3 points) Make a plot that shows the change over the past 100 years for the country with the biggest improvement in life expectancy. Be sure to add appropriate aesthetics to make the plot clean and clear. Once you have made the plot, do a little internet searching and see if you can discover what historical event may have contributed to this remarkable change.**  

```{r}
life_exp_long %>% 
  filter(country == "Kuwait", 
         between(year, 1920, 2020)) %>%
  ggplot(aes(x = as.numeric(year), y = life_expectancy_years))+
  geom_line()+
  labs(x = "Year",
       y = "Life Expectancy (years)",
       title = "Life Expectancy in Kuwait over Time (1922-2022)")+
  theme_linedraw()+
  theme(plot.title = element_text(hjust = 0.5))
```
<br/>
<br/>
Beginning around 1946, Kuwait experienced a large boost to their economy due to the country's export of oil as the resource became more valuable worldwide with urbanization. As a result, Kuwait experienced greater urban development and raised public health standards and available services. ["Kuwait's Modernity Between Memory and Forgetting"](https://www.academia.edu/8186917)

## Population Growth

**6. (3 points) Which 5 countries have had the highest population growth over the past 100 years (1920-2020)?**  

```{r}
population_long %>% 
  group_by(country) %>% 
  filter(year == 1920 | year == 2020) %>% 
  summarize(population_growth = max(population) - min(population)) %>% 
  top_n(n = 5) %>% 
  arrange(desc(population_growth))
```
<br/>
<br/>
**7. (4 points) Produce a plot that shows the 5 countries that have had the highest population growth over the past 100 years (1920-2020). Which countries appear to have had exponential growth?**  

```{r}
population_long %>% 
  filter(country == c("India", "China", "Indonesia", "United States", "Pakistan"), between(year, 1920, 2020)) %>% 
  ggplot(aes(x = as.numeric(year), y = population/(10^6), color = country))+
  geom_line()+
  labs(x = "Year",
       y = "Population Growth (Millions of People)",
       caption = "Population growth over time for the top five growing countries from 1920 to 2020",
       color = "Country")+
  theme_linedraw()+
  theme(plot.caption = element_text(color = "gray50", hjust = 0.5))
```
<br/>
<br/>
It appears that China and India have had exponential growth in their population from 1920-2022.

## Income
The units used for income are gross domestic product per person adjusted for differences in purchasing power in international dollars.
<br/>
**8. (4 points) As in the previous questions, which countries have experienced the biggest growth in per person GDP. Show this as a table and then plot the changes for the top 5 countries. With a bit of research, you should be able to explain the dramatic downturns of the wealthiest economies that occurred during the 1980's.**  
<br/>
From Beginning to End
```{r}
economy_long %>% 
  group_by(country) %>% 
  filter(between(year, 1920, 2022)) %>% 
  summarize(gdp_per_capita_change = max(income_per_person_gdppercapita_ppp_inflation_adjusted) - min(income_per_person_gdppercapita_ppp_inflation_adjusted)) %>% 
  arrange(desc(gdp_per_capita_change))
```

```{r}
economy_long %>% 
  group_by(country) %>% 
  filter(between(year, 1920, 2022)) %>% 
  summarize(gdp_per_capita_change = max(income_per_person_gdppercapita_ppp_inflation_adjusted) - min(income_per_person_gdppercapita_ppp_inflation_adjusted)) %>% 
  arrange(desc(gdp_per_capita_change)) %>% 
  top_n(n = 5) %>% 
  ggplot(aes(x = reorder(country, gdp_per_capita_change), y = gdp_per_capita_change))+
  geom_col(fill = "mediumorchid3")+
  labs(x = "Country",
       y = "Change in GDP per Capita",
       caption = "Change in GDP per capita (adjusted for differences in purchasing power in international dollars)
       for the countries among the top five greatest growing GDP per capita from 1920 to 2020")+
  theme_linedraw()+
  theme(plot.caption = element_text(color = "gray50", hjust = 0.5))
```
<br/>
<br/>
Over Time
```{r}
economy_long %>% 
  filter(country == c("Luxembourg", "Kuwait", "Brunei", "Qatar", "United Arab Emirates")) %>% 
  ggplot(aes(x = as.numeric(year), y = income_per_person_gdppercapita_ppp_inflation_adjusted, color = country))+
  geom_line()+
  labs(x = "Year",
       y = "GDP per Capita",
       caption = "Change in GDP per capita (adjusted for differences in purchasing power in international dollars) over time
       for the countries among the top five greatest growing GDP per capita from 1920 to 2020",
       color = "Country")+
  theme_linedraw()+
  theme(plot.caption = element_text(color = "gray50", hjust = 0.5))
```
<br/>
<br/>
In the early 1980s, there was a global recession due to a major disruption to the oil supply (1979 energy crisis) caused by the Iranian Revolution, leading to a drastic increase in oil prices and high inflation. ["Early 1980s recession"](https://en.wikipedia.org/wiki/Early_1980s_recession)
<br/>
<br/>
**9. (3 points) Create three new objects that restrict each data set (life expectancy, population, income) to the years 1920-2020. Hint: I suggest doing this with the long form of your data. Once this is done, merge all three data sets using the code I provide below. You may need to adjust the code depending on how you have named your objects. I called mine `life_expectancy_100`, `population_100`, and `income_100`. For some of you, learning these `joins` will be important for your project.**  

```{r}
population_100 <- population_long %>% 
  filter(between(year, 1920, 2020))
population_100
```

```{r}
life_expectancy_100 <- life_exp_long %>% 
  filter(between(year, 1920, 2020))
life_expectancy_100
```

```{r}
income_100 <- economy_long %>% 
  filter(between(year, 1920, 2020))
income_100
```

```{r}
gapminder_join <- inner_join(life_expectancy_100, population_100, by= c("country", "year"))
gapminder_join <- inner_join(gapminder_join, income_100, by= c("country", "year"))
gapminder_join
```
<br/>
<br/>
**10. (4 points) Use the joined data to perform an analysis of your choice. The analysis should include a comparison between two or more of the variables `life_expectancy`, `population`, or `income.`**  

How has life expectancy from 1920-2020 changed with GDP per capita for countries with the top 5 and bottom 5 growth in GDP per capita from 1920-2020?  

Breakdown
filter: 1920-2020, top and bottom 5 countries by GDP per capita growth
dependent variable, y: life expectancy
independent variable, x: GDP per capita  

Top 5 Countries
```{r}
gapminder_join %>% 
  group_by(country) %>% 
  filter(between(year, 1920, 2020)) %>% 
  summarize(gdp_growth = max(income_per_person_gdppercapita_ppp_inflation_adjusted) - min(income_per_person_gdppercapita_ppp_inflation_adjusted)) %>% 
  top_n(n = 5) %>% 
  arrange(desc(gdp_growth))
```

```{r}
top_5 <- gapminder_join %>% 
  filter(country == c("United Arab Emirates","Qatar","Brunei","Kuwait",
                      "Luxembourg")) %>% 
  select(-population, -year)
top_5
```

```{r}
top_5 %>% 
  ggplot(aes(x = income_per_person_gdppercapita_ppp_inflation_adjusted, life_expectancy_years))+
  geom_point()+
  geom_smooth(method = "lm")+
  scale_x_log10()+
  labs(x = "Log 10 GDP per Capita",
       y = "Life Expectancy (years)",
       caption = "Life expectancy by log 10 GDP per capita for the top five fastest growing countries by 
       GDP per capita from 1920 to 2020")+
  theme_linedraw()+
  theme(plot.caption = element_text(color = "gray50", hjust = 0.5))
```
<br/>
<br/>
Bottom 5 Countries
```{r}
gapminder_join %>% 
  group_by(country) %>% 
  filter(between(year, 1920, 2020)) %>% 
  summarize(gdp_growth = max(income_per_person_gdppercapita_ppp_inflation_adjusted) - min(income_per_person_gdppercapita_ppp_inflation_adjusted)) %>% 
arrange(gdp_growth) %>% 
  head(n = 5)
```

```{r}
bottom_5 <- gapminder_join %>% 
  filter(country == c("Burundi","Somalia","Niger","Malawi",
                      "Sierra Leone")) %>% 
  select(-population, -year)
bottom_5
```

```{r}
bottom_5 %>% 
  ggplot(aes(x = income_per_person_gdppercapita_ppp_inflation_adjusted, y = life_expectancy_years))+
  geom_point()+
  geom_smooth(method = "lm")+
  scale_x_log10()+
  labs(x = "Log 10 GDP per Capita",
       y = "Life Expectancy (years)",
       caption = "Life expectancy by log 10 GDP per capita for the top five slowest growing countries
       by GDP per capita from 1920 to 2020")+
  theme_linedraw()+
  theme(plot.caption = element_text(hjust = 0.5, color = "gray50"))
```
<br/>
<br/>
Comparison Between Top and Bottom 5's
```{r}
top_5 %>% 
  ggplot(aes(x = income_per_person_gdppercapita_ppp_inflation_adjusted, y = life_expectancy_years, color = "Top Five"))+
  geom_point()+
  geom_point(data = bottom_5, aes(x = income_per_person_gdppercapita_ppp_inflation_adjusted, y = life_expectancy_years, color = "Bottom Five"))+
  scale_x_log10()+
  geom_smooth(method = "lm", se = F)+
  geom_smooth(data = bottom_5, aes(x = income_per_person_gdppercapita_ppp_inflation_adjusted, y = life_expectancy_years, color = "Bottom Five"), method = "lm", se = F)+
  labs(x = "Log 10 GDP per Capita",
       y = "Life Expectancy (years)",
       caption = "Life expectancy by log 10 GDP per capita for the top five fastest 
       and top five slowest growing countries GDP per capita from 1920 to 2020",
       color = "GDP per Capita
       Growth")+
  theme_linedraw()+
  theme(plot.caption = element_text(hjust = 0.5, color = "gray50"))
```
<br/>
<br/>
It seems that life expectancy scales with GDP per capita regardless of how much absolute GDP per capita is amassed since the slopes of both the bottom five and top five fastest growing countries appear to be approximately the same. However it should be noted that the data for the bottom five fastest growing countries are more spread out/inconsistent than that of the top five countries where there is a stronger correlation.
