---
title: "Lab 7 Homework"
author: "Benjamin Prieto"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
``` 

## Data
**1. For this homework, we will use two different data sets. Please load `amniota` and `amphibio`.**  

`amniota` data:  
Myhrvold N, Baldridge E, Chan B, Sivam D, Freeman DL, Ernest SKM (2015). “An amniote life-history
database to perform comparative analyses with birds, mammals, and reptiles.” _Ecology_, *96*, 3109.
doi: 10.1890/15-0846.1 (URL: https://doi.org/10.1890/15-0846.1).
```{r}
amniota <- readr::read_csv("data/amniota.csv")
```

`amphibio` data:  
Oliveira BF, São-Pedro VA, Santos-Barrera G, Penone C, Costa GC (2017). “AmphiBIO, a global database
for amphibian ecological traits.” _Scientific Data_, *4*, 170123. doi: 10.1038/sdata.2017.123 (URL:
https://doi.org/10.1038/sdata.2017.123).
```{r}
amphibio <- readr::read_csv("data/amphibio.csv")
```

## Questions  
**2. Do some exploratory analysis of the `amniota` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  
```{r}
glimpse(amniota)
```

```{r}
summary(amniota)
```


**3. Do some exploratory analysis of the `amphibio` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  
```{r}
glimpse(amphibio)
```

```{r}
summary(amphibio)
```


**4. How many total NA's are in each data set? Do these values make sense? Are NA's represented by values?**   

`amniota`  
```{r}
sum(is.na(amniota))
```

According to `is.na()`, `amniota` does not have any NAs, which is nonsensical. Their NAs are represented by "-999" with varying numbers of decimal points according to each variable's significant figures, as shown by the cursory preview of the data. I'm unsure if that -30258.71100 value is supposed to also signify NAs for `female_maturity_d`. -30258.71100 does not make sense for that variable since it is a time-based variable so it could represent an NA, but because that variable also has -999's that would mean that the dataset has two values for NAs which would be inconsistent. I chose to treat them as NA's since the value is nonsensical for that variable.

`amphibio`  
```{r}
sum(is.na(amphibio))
```

According to `is.na()`, `amphibio` has 170,566 NAs.


**5. Make any necessary replacements in the data such that all NA's appear as "NA".**   
```{r}
amniota_new <- amniota %>% 
  na_if(-999) %>% 
  na_if(-30258.71100)
```

```{r}
summary(amniota_new) #check results
```


**6. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amniota` data.**  
```{r}
naniar::miss_var_summary(amniota_new)
```

**7. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amphibio` data.**
```{r}
naniar::miss_var_summary(amphibio)
```

**8. For the `amniota` data, calculate the number of NAs in the `egg_mass_g` column sorted by taxonomic class; i.e. how many NA's are present in the `egg_mass_g` column in birds, mammals, and reptiles? Does this results make sense biologically? How do these results affect your interpretation of NA's?**  
```{r}
amniota_new %>% 
  group_by(class) %>% 
  select(class, egg_mass_g) %>% 
  naniar::miss_var_summary()
```

Yes, these results makes sense biologically—mammals do not lay eggs (except for platypuses as far as I know) so they do not have any data for `egg_mass_g`, though I am surprised by the number of NAs for birds since I'd expect there to be much more data for them. Perhaps rather than a lack of potential data, birds egg data is just harder to procure due factors such as egg fragility, defensive behavior from parents protecting eggs, difficulty finding eggs hidden from view, extensive predation on these eggs, short incubation period, etc. So I suppose NAs can both represent data that cannot be obtained due to limitations of the subjects themselves (e.g. mammals largely incapable of laying eggs), and data that cannot be obtained due to methodology constraints (e.g. unable to procure samples to measure, measuring tool limitations).


**9. The `amphibio` data have variables that classify species as fossorial (burrowing), terrestrial, aquatic, or arboreal.Calculate the number of NA's in each of these variables. Do you think that the authors intend us to think that there are NA's in these columns or could they represent something else? Explain.**
```{r}
names(amphibio) #check how variables are represented spelling-wise
```

```{r}
amphibio %>% 
  select(Fos,Ter,Aqu,Arb) %>% 
  summarize_all(~(sum(is.na(.))))
```

I do not think they intended to use NAs to represent actual NAs. Based on these variables and their entries, it looks like they used "1" to represent whether each variable was TRUE for each entry and "NA" to represent if the variable was FALSE. Instead they could represent "0" as FALSE instead to avoid confusion with actual missing data, or alternatively use TRUE and FALSE, or some equivalent such as T and F.


**10. Now that we know how NA's are represented in the `amniota` data, how would you load the data such that the values which represent NA's are automatically converted?**
```{r}
amniota_new2 <- readr::read_csv(file = "data/amniota.csv",na =c("-999","-30258.711"))
```

```{r}
glimpse(amniota_new2) #check results
```

```{r}
summary(amniota_new2) #check results
```


## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.  